---
title: "Divvy Tripdata Database"
output: 
  html_document:
    theme:
      bg: "#101010"
      fg: "#FDF7F7" 
      primary: "#3A97D3"
      base_font:
        google: Montserrat
      code_font:
        google: JetBrains Mono
    highlight: espresso
    css: "resources/styles.css"
    toc: true
    toc_depth: 4
    toc_float:
      collapsed: true
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

#Install the following libraries if required
if(!requireNamespace("RPostgres", quietly = TRUE)){
  # If not installed, then install it
  install.packages("RPostgres")
}
if(!requireNamespace("DBI", quietly = TRUE)){
  # If not installed, then install it
  install.packages("DBI")
}
if(!requireNamespace("ini", quietly = TRUE)){
  # If not installed, then install it
  install.packages("ini")
}

# Install necessary libraries
library(RPostgres)
library(DBI)
library(ini)
library(readr)
library(glue)
```

## Introduction



## Divvy Database Creation

### Database connection
Establish postgresql database connection
```{r db_connection}
# Read config
config <- read.ini("resources/db_config.ini")
db <- config$postgresql

# Safe database connection
tryCatch({
  con <- dbConnect(
    Postgres(),
    host = db$host,
    dbname = db$database,
    user = db$user,
    password = db$password,
    port = as.integer(db$port)
  )
}, error = function(e) {
  stop("Database connection failed: ", e$message)
})

# Register connection for SQL chunks
knitr::opts_chunk$set(connection = con)
```

### Database schema
Create database schema to the database
```{sql create_schema}
CREATE SCHEMA IF NOT EXISTS divvy;
```

### Database tables
Read data into R environment
```{r read_data}
# Define year and months
year <- "2024"
months <- sprintf("%02d", 1:12)

# Initialize list to store data
divvy_data <- list()

# Loop through months and read each file
for (m in months) {
  file_path <- paste0("resources/data/", year, m, "-divvy-tripdata.csv")
  month_name <- tolower(format(as.Date(paste0(year, "-", m, "-01")), "%B"))
  
  divvy_data[[month_name]] <- read_csv(file_path, show_col_types = FALSE)
}
```

Drop tables if already exists
```{r drop_monthly_tables, message=FALSE}
# Define all month names
months <- tolower(month.name)

for (m in months) {
  sql <- glue::glue("DROP TABLE IF EXISTS divvy.{m};")
  DBI::dbExecute(con, sql)
}
```

#### January table
Create and load data for the month of January
```{r january_table, message=FALSE}
# Create table
create_jan_table <- "
  CREATE TABLE divvy.january (
      ride_id             TEXT PRIMARY KEY,
      rideable_type       TEXT,
      started_at          TIMESTAMP,
      ended_at            TIMESTAMP,
      start_station_name  TEXT,
      start_station_id    TEXT,
      end_station_name    TEXT,
      end_station_id      TEXT,
      member_casual       TEXT
  );
"
DBI::dbExecute(con, create_jan_table)

# Write to PostgreSQL (replaces table if exists)
dbWriteTable(
  conn      = con,
  name      = DBI::Id(schema = "divvy", table = "january"),
  value     = divvy_data$january,
  overwrite = TRUE,   # drop & recreate table
  row.names = FALSE
)
```

#### February table
Create and load data for the month of February
```{r february_table, message=FALSE}
# Create table
create_feb_table <- "
  CREATE TABLE divvy.february (
      ride_id             TEXT PRIMARY KEY,
      rideable_type       TEXT,
      started_at          TIMESTAMP,
      ended_at            TIMESTAMP,
      start_station_name  TEXT,
      start_station_id    TEXT,
      end_station_name    TEXT,
      end_station_id      TEXT,
      member_casual       TEXT
  );
"
DBI::dbExecute(con, create_feb_table)

# Write to PostgreSQL (replaces table if exists)
dbWriteTable(
  conn      = con,
  name      = DBI::Id(schema = "divvy", table = "february"),
  value     = divvy_data$february,
  overwrite = TRUE,   # drop & recreate table
  row.names = FALSE
)
```

#### March table
Create and load data for the month of January
```{r january_table, message=FALSE}
# Create table
create_jan_table <- "
  CREATE TABLE divvy.january (
      ride_id             TEXT PRIMARY KEY,
      rideable_type       TEXT,
      started_at          TIMESTAMP,
      ended_at            TIMESTAMP,
      start_station_name  TEXT,
      start_station_id    TEXT,
      end_station_name    TEXT,
      end_station_id      TEXT,
      member_casual       TEXT
  );
"
DBI::dbExecute(con, create_jan_table)

# Write to PostgreSQL (replaces table if exists)
dbWriteTable(
  conn      = con,
  name      = DBI::Id(schema = "divvy", table = "january"),
  value     = divvy_data$january,
  overwrite = TRUE,   # drop & recreate table
  row.names = FALSE
)
```

#### April table
Create and load data for the month of April
```{r apr_table, message=FALSE}
# Create table
create_apr_table <- "
  CREATE TABLE divvy.april (
      ride_id             TEXT PRIMARY KEY,
      rideable_type       TEXT,
      started_at          TIMESTAMP,
      ended_at            TIMESTAMP,
      start_station_name  TEXT,
      start_station_id    TEXT,
      end_station_name    TEXT,
      end_station_id      TEXT,
      member_casual       TEXT
  );
"
DBI::dbExecute(con, create_apr_table)

# Write to PostgreSQL (replaces table if exists)
dbWriteTable(
  conn      = con,
  name      = DBI::Id(schema = "divvy", table = "april"),
  value     = divvy_data$april,
  overwrite = TRUE,   # drop & recreate table
  row.names = FALSE
)
```

#### May table
Create and load data for the month of January
```{r january_table, message=FALSE}
# Create table
create_jan_table <- "
  CREATE TABLE divvy.january (
      ride_id             TEXT PRIMARY KEY,
      rideable_type       TEXT,
      started_at          TIMESTAMP,
      ended_at            TIMESTAMP,
      start_station_name  TEXT,
      start_station_id    TEXT,
      end_station_name    TEXT,
      end_station_id      TEXT,
      member_casual       TEXT
  );
"
DBI::dbExecute(con, create_jan_table)

# Write to PostgreSQL (replaces table if exists)
dbWriteTable(
  conn      = con,
  name      = DBI::Id(schema = "divvy", table = "january"),
  value     = divvy_data$january,
  overwrite = TRUE,   # drop & recreate table
  row.names = FALSE
)
```

#### June table
Create and load data for the month of June
```{r jun_table, message=FALSE}
# Create table
create_feb_table <- "
  CREATE TABLE divvy.june (
      ride_id             TEXT PRIMARY KEY,
      rideable_type       TEXT,
      started_at          TIMESTAMP,
      ended_at            TIMESTAMP,
      start_station_name  TEXT,
      start_station_id    TEXT,
      end_station_name    TEXT,
      end_station_id      TEXT,
      member_casual       TEXT
  );
"
DBI::dbExecute(con, create_jun_table)

# Write to PostgreSQL (replaces table if exists)
dbWriteTable(
  conn      = con,
  name      = DBI::Id(schema = "divvy", table = "june"),
  value     = divvy_data$june,
  overwrite = TRUE,   # drop & recreate table
  row.names = FALSE
)
```

















### Database views
```{sql}
SELECT COUNT(*) FROM divvy.january;
```
```{sql}
SELECT COUNT(*) FROM divvy.february;
```

### Database functions



## Exploratory Data Analysis



## Conclusion





```{r disconnect, include=FALSE}
dbDisconnect(con)
```